---
title: "Final report"
author: David Earn and Kevin Zhao
output: html_document
---

This is the required final report for PHAC contract No.~4500489267 ("Re-implementing two versions of the EPAC model").
As specified in Annex Section 3.1 on page 8 of the contract, this report constitutes deliverable 2.

This document was compiled on
```{r echo=FALSE}
Sys.time()
```

In order to reproduce this document from source, including generating all the
files it depends on, please follow the following procedure:

The file `expected_outputs.RData` contains the results of simulations
of the old EPAC model, which uses legacy `macpan2`.  Regenerating these
simulations would require installing the old version and then running
the code, as follows:

```{r eval=FALSE}
## Install old version of EPACmodel:
remotes::install_github("phac-modelling-hub/EPACmodel@v1.3.0")

## load the old EPACmodel
library(EPACmodel)

values = get_default_values("hosp")
values_contact <- values
values_contact$setting.weight <- c(school = 1, work = 1, household = 1, community = 1)
values$transmissibility <- values$transmissibility*(1-0.2) # reduce transmissibility by 20%

old_oy <- make_simulator("old-and-young")
old_oy_change <- make_simulator("old-and-young", "change-transmission")
old_five <- make_simulator("five-year-age-groups")
old_five_change <- make_simulator("five-year-age-groups", "change-contacts")
old_hosp <- make_simulator("hosp")
old_hosp_change <- make_simulator("hosp", "change-contacts")

expected_oy_output <- simulate(old_oy)
expected_oy_output_change <- simulate(old_oy_change)
expected_five_output <- simulate(old_five)
expected_five_output_change <- simulate(old_five_change)
expected_hosp_output <- simulate(old_hosp)
expected_hosp_output_change <- simulate(old_hosp_change)
expected_hosp_mod <- simulate(make_simulator("hosp", values = list(time.steps = 365)), values = values)
expected_hosp_contact_mod <- simulate(make_simulator("hosp", values = list(time.steps = 365)), values = values_contact)

save(
  expected_oy_output, 
  expected_oy_output_change, 
  expected_five_output, 
  expected_five_output_change, 
  expected_hosp_output, 
  expected_hosp_output_change, 
  expected_hosp_mod,
  expected_hosp_contact_mod,
  file = "dev/expected_outputs.RData"
)
```

Whether you have regenerated `expected_outputs.RData` or are just using
the saved version on the `EPACmodel` repo, your next step
is to ensure that you are on the up-to-date fork of `EPACmodel`.  Do this once:
```{r eval=FALSE}
remotes::install_github("KevinHZhao/EPACmodel@217a3e8")
```

# a. Reimplementation progress (model features re-implemented successfully, model features outstanding)

The `hosp` and `five-year-age-groups` models have been succesfully reimplemented
in modern `macpan2`, including the ability to update `hosp` model simulators in
the `simulate()` function.  Side-by-side comparisons of simulations are shown in
section f.

Re-implementation of the `old-and-young` and `hosp-treat` models have also been
completed and included.

# b. New unit tests written and current test coverage

Previous unit tests using the "old-and-young" model have been succesfully 
re-implemented and pass all previous checks (accounting for differences in
structure from our re-implementation).
Unit tests equivalent to these "old-and-young" model tests have also been added for
the `get_model()` and `simulate()` functions when using the "five-year-age-groups" model.
For the `make_simulator()` tests, one of the tests checks that "transmission" is
properly updated to be all "0" when given the appropriate `values`.
In the previous version of `EPACmodel` using the old version of `macpan2`,
"transmission" was obtained by running

```{r eval=FALSE}
    make_simulator(
      model.name = "five-year-age-groups",
      values = values)$tmb_model$data_arg()$mats[[3]]
```

but due to re-ordering in our updates, we have modified this test 
to refer to `[[1]]` rather `[[3]]`, so we run

```{r eval=FALSE}
    make_simulator(
      model.name = "five-year-age-groups",
      values = values)$tmb_model$data_arg()$mats[[1]]
```

to properly access "transmission".

We have added simple snapshot tests for the default "five-year-age-groups",
"hosp", and "old-and-young" models in all three of the above functions, as well
as the "change-contacts" scenario for the `make_simulator()` and `simulate()`
functions.  Many of the simulation comparisons in section e have been
formally implemented as tests as well, using the RData file generated at the
top of this script for comparison.  These tests are all for the `simulate()`
function, and include:

- ensuring that modifying the initial state, the contact parameters, and flow
parameters in `simulate` produce correct results.
- ensuring that the "change-contacts" scenario produces identical state values
at each time step as the old implementation of `EPACmodel`, in the "hosp" and
"five-year-age-groups" models.
- ensuring that the "change-transmission" scenario produces identical state values
at each time step as the old implementation of `EPACmodel`, in the "old-and-young"
model.

A snapshot test has also been added for the new `list_flows()` function.
Some tests produce warnings due to either an intended warning when updating
values in the `simulate()` function, warning the user that only the `hosp` model
accepts these modifications, and another when the user tries to modify initial state variables
in the `simulate()` function, seemingly due to how TMB views the parameter frame.
These warnings do not appear to be indicative of any issues, and can likely be
suppressed in the package.

# c. New package documentation added

- `?get_model` has been modified to specify that the returned object is of class 
`TMBModelSpec` instead of `Model`.
- `?list_flows` has been added, as well as information in each model's README
file specifying what each of these flows in each model represent.
- `?make_simulator` now has an "output_flows" argument, which is a vector of flows
from `list_flows()` that the user wishes the model would return when simulating.
- `?simulate` has been modified to specify that the output is now a five column
data frame, where age and variable names are now separate, and the `value_type`
column is now only ever "state" or "flow" ("total_inflows" from the previous
implementation are now gone, but equivalent information can be obtained from flow
variables).
- The `EPACmodel.Rmd` and `local-models.Rmd` vignettes have been updated 
(code and documentation) to give information about the re-implemented version of
`EPACmodel`.

# d. New GitHub issues opened

No new GitHub issues have had to be opened, as we have not encountered any 
bugs/unexpected behaviours in the package that have not already been discussed 
and/or fixed by the re-implementation.

# e. Simulation results from the completed re-implementation work demonstrating all features of the model

We have updated the `make_simulator()`, `get_model()`, and `simulate()` functions
to use modern `macpan2` when creating and simulating each model in the `EPACmodel`
package, including in each model's available scenarios ("change-contacts" for 
`hosp`, `hosp-treat` and `five-year-age-groups`/"change-transmission" for
`old-and-young`).

In the current version of `macpan2`, `macpan2::Model` objects no
longer exist, and `get_model()` now produces `TMB::TMBModelSpec`
objects.  We revised `get_model()` to make the creation of age
structures easier; the default values of the state variables and flow
parameters are now scalars that ignore age structure, rather than
scalars for each of the age grouped state variables and flow
parameters.  These default values can be modified when the simulator
object is built (with `make_simulator()`) to be vectors, as we will
later show.  We also simplified the model expressions, as they use the
age grouped state variable vectors and flow parameter vectors (which
are stored at the end of each time step if they are updated) to
compute the flows at each time step.  An advantage of these
modifications is that now, when building the simulator object, it is
very simple to modify the code with different age groups by modifying
`age.group.lower` in the `run-before-simulator.R` script or making
`age.group.lower` a user specified argument when calling
`make_simulator`, and supplying the relevant `values` appropriate to
the age groupings.

We can illustrate structural changes to the returned model, simulator and simulations
by simulating the "hosp" model:

```{r}
library(EPACmodel)
library(dplyr)
library(waldo)

hosp_model <- get_model("hosp")
print(hosp_model)

hosp_sim <- make_simulator("hosp")
hosp_output <- simulate(hosp_sim)
print(head(hosp_output))
```
`simulate` now returns more "tidy" data, where instead of outputting variables
in a format similar to "S.lb0", the age group and variable name are now their
own columns.  "variable_name" is returned as a factor so that the ordering
of this column is preserved.  "value_type" is now either "state" or "flow".  
By default, only state variables are returned by simulate, but we may use the `list_flows()`
function to get a list of returnable flows and the "output_flows" argument of
`make_simulator` to choose which flows we wish to return.  For example:

```{r}
list_flows("hosp")

hosp_sim <- make_simulator("hosp", output_flows = c("prog_R", "prog_A", "prog_D"))
hosp_output <- simulate(hosp_sim)
print(head(hosp_output |> filter(value_type == "flow")))
```

In the above example, there are rows for the values of each `prog_x` flow, where
`x` is `R`, `A`, or `D`.  These represent the number of people progressing from
the `E` compartment to the various `I_x` compartments at each time step, giving
us the model's predictions for how many individuals become infectious at each
time step.

Simulations with the "change-contacts" scenario have also been succesfully
re-implemented, and parameter values can be modified in the `simulate` function
without having to regenerate the simulator object:

```{r}
  simulator = make_simulator(model.name = "hosp", scenario.name = "change-contacts")
  values = get_default_values("hosp")
  values_state <- values
  values_state$state["S.lb10"] <- 0
  hosp_mod <- simulate(make_simulator("hosp"), values = values_state)
  S.lb10 <- hosp_mod |> dplyr::filter(variable_name == "S", age_group == 10) |> dplyr::pull(value)
  
  all(S.lb10 == 0)
```

As expected, when we adjust the initial value of `S` in the 10-15 year old age group
to 0, it remains 0 for all time.  Note that a warning is output, which appears to
occur always when a state variable is modified in the simulate function.  This
appears to be on the TMB side when modifying the parameter frame, but it does not
appear to cause any issues with the output and can likely be safely ignored.

In the old implementation, all parameters were added to the parameter frame when
the simulator was built, so that TMB could run simulations with modified parameter
frames.  We have modified it now so that the parameter frame is only modified when
trying to replace parameters via `simulate()`.

Simulating for the "five-year-age-groups" model also works in the default case
and in the "change-contacts" scenario.

```{r}
five_sim <- make_simulator("five-year-age-groups")
five_output <- simulate(five_sim)
head(five_output)
tail(five_output)
```

Simulations using the `hosp-treat` and `old-and-young` models also work.  For
brevity we will only show the `old-and-young` model in the "change-transmission"
scenario:

```{r}
oy_sim <- make_simulator("old-and-young", "change-transmission")
oy_output <- simulate(oy_sim)
head(oy_output)
tail(oy_output)
```

# f.	A side-by-side comparison of simulation results from the old and new implementations

Our updates to the simulator functions still produce results consistent with
the old version of `EPACmodel`.  Automated tests to confirm this for the
`old-and-young`, `five-year-age-groups` and `hosp` models have been created.
We will demonstrate some of these tests for side-by-side comparison.

First we note some major changes we have made from the old `EPACmodel` structure.
In the previous version of `EPACmodel`, the simulation
would return an extra time step (ending on the 61st time step instead of the
60th by default).
This appears to not be an intentional feature, and in our updated version, this
extra time step is no longer present.

```{r}
load("expected_outputs.RData")
unique(expected_five_output$time)
```

As noted before, the `total_inflow` rows are no longer returned, hence for our
comparison we will only examine the returned state variables.

We can verify that the returned `value`'s are near identical matches by 
formatting the old `EPACmodel` output to match our current structure.
For instance, let us compare a `hosp` simulation in the "change-contacts" scenario
where contact parameters are modified:

```{r}
simulator = make_simulator(model.name = "hosp", scenario.name = "change-contacts")
values = get_default_values("hosp")
values_contact <- values
values_contact$setting.weight <- c(school = 1, work = 1, household = 1, community = 1)
hosp_mod <- simulate(make_simulator("hosp", values = list(time.steps = 365)), values = values_contact) |>
  dplyr::filter(value_type == "state")

## For formating old structure
tidy_output = function(output, levels, value_type = "state") {
  if (!is.null(value_type)) {
    output <- (output
               |> dplyr::filter(value_type %in% !!value_type)
    )
  }
  output <- (output
             # parse state names
             |> tidyr::separate(
               state_name,
               into = c("var", "age"),
               sep = ".lb"
             )
             # enforce order of states
             |> dplyr::mutate(
               var = factor(var, levels = levels)
             )
  )
  output
}

expected_hosp <- tidy_output(expected_hosp_contact_mod, levels = levels(hosp_mod$variable_name)) |>
  rename(age_group = age, variable_name = var) |>
  select(time, age_group, variable_name, value_type, value) |>
  filter(value_type == "state", time != 366) |>
  mutate(age_group = as.numeric(age_group)) |>
  arrange(time, variable_name, age_group)

compare(expected_hosp, hosp_mod, tolerance = 1e-12)
```

Comparing "five-year-age-groups" with "change-contacts":

```{r}
simulator = make_simulator(model.name = "five-year-age-groups", scenario.name = "change-contacts")
five <- simulate(simulator) |>
  dplyr::filter(value_type == "state")

expected_five <- tidy_output(expected_five_output_change, levels = levels(five$variable_name)) |>
  rename(age_group = age, variable_name = var) |>
  select(time, age_group, variable_name, value_type, value) |>
  filter(value_type == "state", time != 61) |>
  mutate(age_group = as.numeric(age_group)) |>
  arrange(time, variable_name, age_group)

compare(expected_five, five, tolerance = 1e-12)
```

Comparing "old-and-young" with "change-transmission"
```{r}
tidy_output_yo = function(output, levels){
  output <- (output
             # parse state names
             |> tidyr::separate(
               state_name,
               into = c("var", "age"),
               sep = "_"
             )
             # enforce order of states
             |> dplyr::mutate(
               var = factor(var, levels)
             )
  )
  output
}

expected_oy <- tidy_output_yo(expected_oy_output_change, levels = levels(oy_output$variable_name)) |>
  dplyr::rename(age_group = age, variable_name = var) |>
  dplyr::select(time, age_group, variable_name, value_type, value) |>
  dplyr::filter(value_type == "state", time != 151) |>
  arrange(time, variable_name, age_group)

compare(oy_output |> arrange(time, variable_name, age_group), expected_oy, tolerance = 1e-12)
```
